---
title: "TCD ALS Immunology Analysis"
author:
  name: "Alejandro Caravaca Puchades"
  email: "acaravaca@idibell.cat"
  affiliation: "Institut d'Investigació Biomèdica de Bellvitge (IDIBELL), L'Hospitalet de Llobregat, Barcelona, Spain"
  orcid: "0000-0003-2223-7335"
toc: true
toc_depth: 2
toc_float: true
df-print: paged
format:
  html:
    embed-resources: true
---

```{r}
#| echo: false
#| message: false

library(boot)
library(broom)
library(ggsignif)
library(gtsummary)
library(ggplot2)
library(ggeffects)
library(ggsurvfit)
library(mice)
library(Rtsne)
library(survival)
library(tidyverse)

set.seed(9999)
```

# Loading the dataset

```{r}
#| echo: false
#| message: false
source("R/data.r")
```

```{r}
imm_pheno
```

```{r}
imm_cytokines
```

```{r}
imm_treg_tmem
```

# Immunophenotype panels

```{r}
imm_pheno.tsne.in <- imm_pheno.imputed |> select(sample_id, group, ends_with("T0"))
imm_pheno.tsne <- Rtsne(imm_pheno.tsne.in |> select(-sample_id, -group), perplexity = 15)
tibble(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y)) +
  geom_point()
```

## Sex

```{r}
imm_pheno.tsne.in |>
  bind_cols(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(clinical |> select(sample_id, sex), by = "sample_id") |>
  ggplot(aes(tSNE.x, tSNE.y, color = sex)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Cases vs controls

```{r}
imm_pheno.tsne.in |>
  bind_cols(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y, color = group)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Site of onset

```{r}
imm_pheno.tsne.in |>
  bind_cols(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, site_of_onset),
    by = "sample_id"
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = site_of_onset)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Slow vs fast progressors

```{r}
imm_pheno.tsne.in |>
  bind_cols(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, progression_category),
    by = "sample_id"
  ) |>
  drop_na(progression_category) |>
  ggplot(aes(tSNE.x, tSNE.y, color = progression_category)) +
  geom_point() +
  theme(legend.position = "bottom")
```

# Treg/Tmem panels

```{r}
imm_treg_tmem.tsne.in <- imm_treg_tmem.imputed |>
  select(sample_id, group, ends_with("T0"))
imm_treg_tmem.tsne <- Rtsne(
  imm_treg_tmem.tsne.in |> select(where(is.numeric)),
  perplexity = 15
)
tibble(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y)) +
  geom_point()
```

## Sex

```{r}
imm_treg_tmem.tsne.in |>
  bind_cols(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(clinical |> select(sample_id, sex), by = "sample_id") |>
  ggplot(aes(tSNE.x, tSNE.y, color = sex)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Cases vs controls

```{r}
imm_treg_tmem.tsne.in |>
  bind_cols(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y, color = group)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Site of onset

```{r}
imm_treg_tmem.tsne.in |>
  bind_cols(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, site_of_onset),
    by = "sample_id"
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = site_of_onset)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Slow vs fast progressors

```{r}
imm_treg_tmem.tsne.in |>
  bind_cols(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  left_join(
    clinical |> select(sample_id, progression_category),
    by = "sample_id"
  ) |>
  filter(group == "ALS") |>
  drop_na(progression_category) |>
  ggplot(aes(tSNE.x, tSNE.y, color = progression_category)) +
  geom_point() +
  theme(legend.position = "bottom")
```

# CD4 cytokine expression

```{r}
imm_cyto_cd4.tsne.in <- imm_cytokines_l.imputed |>
  filter(celltype == "CD4", timepoint == "T0") |>
  select(-celltype, -timepoint)
imm_cyto_cd4.tsne <- Rtsne(imm_cyto_cd4.tsne.in |> select(-sample_id, -group), perplexity = 15)
tibble(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y)) +
  geom_point()
```

## Sex

```{r}
imm_cyto_cd4.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(clinical |> select(sample_id, sex), by = "sample_id") |>
  ggplot(aes(tSNE.x, tSNE.y, color = sex)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Cases vs controls

```{r}
imm_cyto_cd4.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y, color = group)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Site of onset

```{r}
imm_cyto_cd4.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, site_of_onset),
    by = "sample_id"
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = site_of_onset)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Slow vs fast progressors

```{r}
imm_cyto_cd4.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, progression_category),
    by = "sample_id"
  ) |>
  drop_na(progression_category) |>
  ggplot(aes(tSNE.x, tSNE.y, color = progression_category)) +
  geom_point() +
  theme(legend.position = "bottom")
```

# CD8 cytokine expression

```{r}
imm_cyto_cd8.tsne.in <- imm_cytokines_l.imputed |>
  filter(celltype == "CD8", timepoint == "T0") |>
  select(-celltype, -timepoint, -ends_with("_tregs"))

imm_cyto_cd8.tsne <- Rtsne(imm_cyto_cd8.tsne.in |> select(-sample_id, -group), perplexity = 15)
tibble(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y)) +
  geom_point()
```

## Cluster analysis

```{r}
imm_cyto_cd8.kmc <- kmeans(
  imm_cyto_cd8.tsne.in |> select(where(is.numeric)),
  centers = 3, nstart = 10
)

imm_cyto_cd8.tsne.in |>
  bind_cols(
    tSNE.x = imm_cyto_cd8.tsne$Y[, 1],
    tSNE.y = imm_cyto_cd8.tsne$Y[, 2],
    KM = factor(imm_cyto_cd8.kmc$cluster)
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = KM)) +
  geom_point() +
  theme(legend.position = "bottom")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(clinical, by = "sample_id") |>
  tbl_summary(by = "KM", include = c(
    sex, age_at_onset, site_of_onset, delta_fs, progression_category
  )) |>
  add_p() |>
  add_significance_stars() |>
  modify_header(label ~ "**CD8-related cluster**")
```

Interestingly, there seems to be a predominance of males in CD8-related cluster 3, as well as a tendency towards a higher delta-FS in CD8-related cluster 2.

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  relocate(KM, .after = group) |>
  filter(KM == 2)
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  relocate(KM, .after = group) |>
  filter(KM == 3)
```

## Principal component analysis

```{r}
imm_cyto_cd8.pca <- prcomp(
  imm_cyto_cd8.tsne.in |> select(-sample_id, -group),
  center = TRUE, scale. = TRUE)
```

```{r}
tibble(eigenval = imm_cyto_cd8.pca$sdev^2) |>
  mutate(PC = row_number()) |>
  ggplot(aes(PC, eigenval / sum(eigenval))) +
  geom_point() + geom_line() +
  labs(x = "Principal component", y = "Eigenvalue") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal()
```

We can see that the first 5 PCs explain around `r round(sum(imm_cyto_cd8.pca$sdev[1:5]^2) / sum(imm_cyto_cd8.pca$sdev^2) * 100, 1)`% of the variance in the data.

```{r}
imm_cyto_cd8.pca$rotation[, 1:5] |>
  as.data.frame() |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "PC", values_to = "loading") |>
  ggplot(aes(PC, marker, fill = loading)) +
  geom_tile() +
  geom_text(aes(label = round(loading, 2)), size = 3) +
  scale_fill_gradient2(low="blue", mid="white", high="red") +
  labs(x = NULL, y = NULL) +
  theme_minimal()
```

This is how the first 5 PCs are driven by the different cytokine markers:

-   **PC1**: driven positively by IL-2, IL-4, GM-CSF and TNF-alpha.
-   **PC2**: driven positively by IL-17 and IL-22.
-   **PC3**: driven negatively by IFN-gamma, IL-21, CD-161 and TNF-alpha.
-   **PC4**: driven positively by CD-161, IL-17 and IL-22, and negatively by IFN-gamma.
-   **PC5**: driven negatively by CD-161 and negatively by IL-21 and IL-22.

## Correlation of CD8 cytokines with age

Now let's explore the correlation between the different PCs and the age of the subjects:

```{r}
imm_cyto_cd8.tsne.in |>
  select(sample_id) |>
  bind_cols(as_tibble(imm_cyto_cd8.pca$x)) |>
  left_join(demographics |> select(sample_id, age), by = "sample_id") |>
  drop_na(age) |>
  pivot_longer(
    -c(sample_id, age),
    names_to = "marker", values_to = "value"
  ) |>
  reframe(cor.test(age, value, method = "kendall") |> tidy(), .by = marker)
```

```{r}
imm_cyto_cd8.tsne.in |>
  select(sample_id) |>
  bind_cols(as_tibble(imm_cyto_cd8.pca$x[,1:5])) |>
  left_join(demographics |> select(sample_id, age), by = "sample_id") |>
  drop_na(age) |>
  pivot_longer(
    -c(sample_id, age),
    names_to = "marker", values_to = "value"
  ) |>
  ggplot(aes(age, value)) +
  geom_smooth(method="lm", formula = y ~ x) +
  geom_point() +
  facet_wrap(~marker, scales = "free")
```

**PC3 seems to be the only primary component statistically associated with age**, and it is negatively correlated with it. Now let's take a look at each of the markers individually:

```{r}
imm_cyto_cd8.tsne.in |>
  left_join(demographics |> select(sample_id, age), by = "sample_id") |>
  pivot_longer(-c(sample_id, group, age), names_to = "marker") |>
  reframe(cor.test(value, age) |> tidy(), .by=marker) |>
  arrange(p.value)
```

```{r}
imm_cyto_cd8.tsne.in |>
  select(sample_id, ifn_gamma, il21, cd161, tnf_alpha) |>
  left_join(demographics |> select(sample_id, age), by = "sample_id") |>
  drop_na(age) |>
  pivot_longer(
    -c(sample_id, age),
    names_to = "marker", values_to = "value"
  ) |>
  ggplot(aes(age, value)) +
  geom_smooth(method="lm", formula = y ~ x) +
  geom_point() +
  facet_wrap(~marker, scales = "free")
```

**It seems like the apparent negative correlation between PC3 and age is mostly driven by IFN-gamma and TNF-alpha**, both being positively correlated with age and inversely represented in PC3.

Let's now see how the first 5 PCs are represented among the clusters we identified earlier:

```{r}
imm_cyto_cd8.tsne.in |>
  select(sample_id, group) |>
  bind_cols(KM=imm_cyto_cd8.kmc$cluster) |>
  bind_cols(imm_cyto_cd8.pca$x[,1:5]) |>
  pivot_longer(starts_with("PC"), names_to = "PC", values_to = "value") |>
  ggplot(aes(factor(KM), value, fill=factor(KM))) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("1", "3"), c("1", "2"), c("2", "3")),
    map_signif_level = TRUE,
    textsize = 3,
    size = 0.5,
    y_position = c(8.5, 7, 7)
  ) +
  coord_cartesian(ylim=c(-5, 10)) +
  facet_wrap(~PC) +
  labs(title = "Imputed dataset")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(cd8_group=factor(imm_cyto_cd8.kmc$cluster)) |>
  pivot_longer(
    -c(sample_id, group, cd8_group),
    names_to = "PC", values_to = "value"
  ) |>
  ggplot(aes(cd8_group, value, fill=cd8_group)) +
  geom_boxplot() +
  facet_wrap(~PC, scales = "free") +
  labs(title = "Imputed dataset")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM=imm_cyto_cd8.kmc$cluster) |>
  pivot_longer(-c(sample_id, group, KM), names_to = "marker") |>
  group_by(marker) |>
  reframe(kruskal.test(value ~ KM) |> tidy()) |>
  arrange(p.value)
```

```{r}
imm_cyto_cd8.tsne.in |>
  select(sample_id) |>
  bind_cols(KM=factor(imm_cyto_cd8.kmc$cluster)) |>
  inner_join(imm_cytokines, by = "sample_id") |>
  drop_na(KM, CD8_il21_T0) |>
  ggplot(aes(KM, CD8_il21_T0, fill=KM)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("1", "3"), c("1", "2"), c("2", "3")),
    map_signif_level = TRUE,
    textsize = 3,
    size = 0.5,
    y_position = c(100, 90, 90)
  ) +
  coord_cartesian(ylim=c(0, 110)) +
  labs(title = "Raw dataset")
```

## Sex

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(clinical |> select(sample_id, sex), by = "sample_id") |>
  ggplot(aes(tSNE.x, tSNE.y, color = sex)) +
  geom_point() +
  theme(legend.position = "bottom")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(clinical |> select(sample_id, sex), by = "sample_id") |>
  drop_na(sex) |>
  ggplot(aes(KM, fill=sex)) +
  geom_bar(position="fill")
```

## Cases vs controls

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y, color = group)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Site of onset

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(clinical |> select(sample_id, site_of_onset), by = "sample_id") |>
  ggplot(aes(tSNE.x, tSNE.y, color = site_of_onset)) +
  geom_point() +
  theme(legend.position = "bottom")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(clinical |> select(sample_id, site_of_onset), by = "sample_id") |>
  drop_na(site_of_onset) |>
  ggplot(aes(KM, fill=site_of_onset)) +
  geom_bar(position="fill")
```

## Slow vs fast progressors

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, progression_category),
    by = "sample_id"
  ) |>
  drop_na(progression_category) |>
  ggplot(aes(tSNE.x, tSNE.y, color = progression_category)) +
  geom_point() +
  theme(legend.position = "bottom")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(
    clinical |> select(sample_id, delta_fs),
    by = "sample_id"
  ) |>
  drop_na(delta_fs) |>
  ggplot(aes(KM, delta_fs, fill=KM)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("1", "3"), c("1", "2"), c("2", "3")),
    map_signif_level = TRUE,
    y_position = c(4, 3.5, 3.5)
  ) +
  coord_cartesian(ylim=c(0, 4.5))
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(clinical, by = "sample_id") |>
  ggplot(aes(il21, delta_fs, color=KM)) +
  geom_point()
```

## Patient survival

```{r}
imm_cyto_cd8.surv <- imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(clinical, by = "sample_id")

imm_cyto_cd8.km <- survfit2(
  Surv(disease_duration, vital_status) ~ KM,
  data = imm_cyto_cd8.surv
)

ggsurvfit(imm_cyto_cd8.km) + add_pvalue() + labs(title="Imputed dataset")
```

There seems to be a **worse survival associated with CD8-related cluster 2**.

```{r}
cd8_km_sex.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM + sex,
  data = imm_cyto_cd8.surv
)

summary(cd8_km_sex.imp_cph)
```

```{r}
cd8_km_sex_i.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM * sex,
  data = imm_cyto_cd8.surv
)

summary(cd8_km_sex_i.imp_cph)
```

This effect is independent of sex, and there's not an associated interaction.

```{r}
cd8_km_onset.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM + site_of_onset,
  data = imm_cyto_cd8.surv
)

summary(cd8_km_onset.imp_cph)
```

```{r}
cd8_km_onset_i.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM * site_of_onset,
  data = imm_cyto_cd8.surv
)

summary(cd8_km_onset_i.imp_cph)
```

This effect is also independent of site of onset, and there's not a clear interaction either.

```{r}
cd8_km_age.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM + age_at_onset,
  data = imm_cyto_cd8.surv
)

summary(cd8_km_age.imp_cph)
```

```{r}
cd8_km_age_i.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM * age_at_onset,
  data = imm_cyto_cd8.surv
)

summary(cd8_km_age_i.imp_cph)
```

Finally, this effect is also not fully explained by differences in age at onset, but there seems to be an interaction where **patients with an earlier age at onset have the worst survival among patients in cluster 2**.

To establish which cytokines among the ones differentially expressed in cluster 2 drive the survival effects, let's analyze each one of them separately:

```{r}
coxph(Surv(disease_duration, vital_status) ~ il21, imm_cyto_cd8.surv)
```

```{r}
coxph(Surv(disease_duration, vital_status) ~ ifn_gamma, imm_cyto_cd8.surv)
```

```{r}
coxph(Surv(disease_duration, vital_status) ~ tnf_alpha, imm_cyto_cd8.surv)
```

```{r}
coxph(Surv(disease_duration, vital_status) ~ gm_csf, imm_cyto_cd8.surv)
```

Among IL-21, IFN-gamma, TNF-alpha and GM-CSF, **IL-21 is the only one that seems to have a significant effect on survival**.

```{r}
coxph(Surv(disease_duration, vital_status) ~ KM + il21, imm_cyto_cd8.surv) |> summary()
```

Adjusting for IL-21 makes the effect of CD8-related cluster 2 non-significant, which suggests that **IL-21 is the main driver of the survival effect we see in this cluster**.

> **Hypothesis**: IL-21 expression in CD8 is the main driver of the survival differences we see in this dataset. Alternatively, it could be inversely-associated with survival if patients with more rapidly progressive ALS or at more advanced stages of disease have higher IL-21 expression by CD8.

Let's test that hypothesis graphically by plotting survival curves of patients above and below the median IL-21:

```{r}
imm_cyto_cd8.surv |>
  mutate(il21_group = if_else(il21 <= median(il21), "IL-21/Low", "IL-21/High")) |>
  with(survfit2(Surv(disease_duration, vital_status) ~ il21_group)) |>
  ggsurvfit() + add_pvalue() + labs(title = "Imputed dataset")
```

We see that **patients in the IL-21/High group have a significantly shorter survival time compared to those in the IL-21/Low group (p = 0.03)**. This suggests that IL-21 is a key factor in determining survival differences in this dataset.

Let's take a look now at the effects of CD8-associated IL-21 on survival, and its interactions with clinical characteristics:

```{r}
cd8_il21_sex_i.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ il21 * sex,
  imm_cyto_cd8.surv
)

summary(cd8_il21_sex_i.imp_cph)
```

```{r}
cd8_il21_onset_i.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ il21 * site_of_onset,
  imm_cyto_cd8.surv
)

summary(cd8_il21_onset_i.imp_cph)
```

```{r}
cd8_il21_age_i.imp_cph <- coxph(
  Surv(disease_duration, vital_status) ~ il21 * age_at_onset,
  imm_cyto_cd8.surv
)

summary(cd8_il21_age_i.imp_cph)
```

Similarly to the survival effects observed for CD8-related cluster 2, **the effects of CD8-related IL-21 is independent of sex, site of onset and age at onset**, and a significant interaction with age at onset is observed.

```{r}
ggpredict(
  cd8_il21_age_i.imp_cph,
  terms = c("il21", "age_at_onset")
) |> plot() + labs(title = "Predicted risk scores – Imputed dataset")
```

## Effects of IL-21 expression at different timepoints

```{r}
imm_cytokines.imputed |>
  filter(group == "ALS") |>
  mutate(il21_group = factor(case_when(
    CD8_il21_T0 > median(CD8_il21_T0, na.rm = TRUE) ~ "IL-21/EarlyHigh",
    CD8_il21_T1 > median(CD8_il21_T1, na.rm = TRUE) ~ "IL-21/LateHigh",
    TRUE ~ "IL-21/Low"
  ), levels = c("IL-21/Low", "IL-21/EarlyHigh", "IL-21/LateHigh"))) |>
  left_join(clinical, by = "sample_id") |>
  with(survfit2(Surv(disease_duration, vital_status) ~ il21_group)) |>
  ggsurvfit() + add_pvalue() + labs(title="Imputed dataset")
```

Patients with **increased CD8 IL-21 expression at both T0 and T1 show a significantly worse survival** compared to those with low IL-21 expression at both timepoints on the imputed dataset (p = 0.039).

Let's visualize the relationship between CD8-related clusters and evolution in IL-21:

```{r}
imm_cytokines_l |>
  filter(group == "ALS", celltype=="CD8") |>
  drop_na(sample_id, timepoint, il21) |>
  left_join(
    imm_cyto_cd8.tsne.in |>
        select(sample_id) |>
        bind_cols(KM=factor(imm_cyto_cd8.kmc$cluster)),
    by = "sample_id"
  ) |>
  mutate(t_pos = case_match(timepoint, "T0" ~ 0, "T1" ~ 1) + rnorm(n(), sd=0.1)) |>
  ggplot(aes(t_pos, il21, group=sample_id, color=KM)) +
  geom_point() +
  geom_line(linetype="dotted") +
  scale_x_continuous(breaks = 0:1, labels = c("T0", "T1")) +
  labs(title = "Raw dataset", x = NULL, y = "CD8 IL-21")
```

## Sensitivity analysis on the unimputed dataset

As we saw previously with CD8-related cluster 2, **the effect of increased IL-21 CD8 response is also associated with worse survival, and this effect is more pronounced in patients with an earlier age at onset**. Let's try to reproduce the same analyses on CD8-associated IL-21 response using the unimputed dataset. This will allow us to assess the robustness of our findings and see if the imputation process has introduced any biases.

```{r}
imm_cytokines_t0 |>
  filter(group == "ALS", timepoint == "T0") |>
  mutate(il21_group = ifelse(
    CD8_il21 <= median(CD8_il21, na.rm = TRUE), "IL-21/Low", "IL-21/High"
  )) |>
  left_join(clinical, by = "sample_id") |>
  with(survfit2(Surv(disease_duration, vital_status) ~ il21_group)) |>
  ggsurvfit() + add_pvalue() + labs(title = "Raw dataset")
```

The same analysis by discretizing on the median IL-21 shows a similar trend in the unimputed dataset, with results not reaching statistical significance. This might be due to both sample size reduction from listwise deletion, and loss of statistical power due to discretization of IL-21 levels.

However, **the Cox regression analysis on the unimputed dataset still shows a highly significant survival effect for IL21-associated CD8 response and a significant interaction with age at onset**, mirroring the effects we saw on the imputed dataset.

```{r}
imm_cytokines |>
  inner_join(clinical, by = "sample_id") |>
  with(coxph(Surv(disease_duration, vital_status) ~ CD8_il21_T0)) |>
  summary()
```

```{r}
cd8_il21_age_i.raw_cph <- imm_cytokines |>
  inner_join(clinical, by = "sample_id") |>
  with(coxph(Surv(disease_duration, vital_status) ~ CD8_il21_T0 * age_at_onset))

summary(cd8_il21_age_i.raw_cph)
```

```{r}
ggpredict(
  cd8_il21_age_i.raw_cph,
  terms = c("CD8_il21_T0", "age_at_onset")
) |> plot() + labs(title = "Predicted risk scores – Raw dataset")
```

## Bootstrapping analysis

To get further confirmation that these results are not the result of overfitting the data due to small sample size, let's get bootstrapped confidence intervals for the Cox regression coefficients:

```{r}
cd8_il21_age_i.boot_cph <- boot(imm_cyto_cd8.surv, function(data, idx) {
	model.fit <- with(data[idx,],
		coxph(Surv(disease_duration, vital_status) ~ il21 * age_at_onset)
	)
	result <- summary(model.fit)
	return(result$coefficients[,c("exp(coef)", "se(coef)")])
}, R = 1000)
```

```{r}
plot(cd8_il21_age_i.boot_cph)
```

```{r}
boot.ci(cd8_il21_age_i.boot_cph)
```

**Bootstrapped confidence intervals for the effect of IL21-based CD8 response on survival all show significant results independently of age at onset**, boosting our confidence in the robustness of these results.