---
title: "TCD ALS Immunology Analysis"
author: "Alejandro Caravaca Puchades"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

# Initialize the environment

```{r echo=FALSE}
library(Rtsne)
library(broom)
library(survival)
library(ggsignif)
library(gtsummary)
library(tidyverse)
library(ggplot2)
library(ggsurvfit)
library(powerSurvEpi)
```

```{r}
set.seed(9999)
```

# Load the dataset

```{r echo=FALSE}
source("R/data.r")
```

```{r}
imm_pheno
```

```{r}
imm_cytokines
```

```{r}
imm_treg_tmem
```

# Immunophenotype panels

```{r}
imm_pheno.tsne.in <- imm_pheno.imputed |> select(sample_id, group, ends_with("T0"))
imm_pheno.tsne <- Rtsne(imm_pheno.tsne.in |> select(-sample_id, -group), perplexity = 15)
tibble(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y)) +
  geom_point()
```

## Cases vs controls

```{r}
imm_pheno.tsne.in |>
  bind_cols(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y, color = group)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Site of onset

```{r}
imm_pheno.tsne.in |>
  bind_cols(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, site_of_onset),
    by = "sample_id"
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = site_of_onset)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Slow vs fast progressors

```{r}
imm_pheno.tsne.in |>
  bind_cols(tSNE.x = imm_pheno.tsne$Y[, 1], tSNE.y = imm_pheno.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, progression_category),
    by = "sample_id"
  ) |>
  drop_na(progression_category) |>
  ggplot(aes(tSNE.x, tSNE.y, color = progression_category)) +
  geom_point() +
  theme(legend.position = "bottom")
```

# Treg/Tmem panels

```{r}
imm_treg_tmem.tsne.in <- imm_treg_tmem.imputed |>
  select(sample_id, group, ends_with("T0"))
imm_treg_tmem.tsne <- Rtsne(
  imm_treg_tmem.tsne.in |> select(where(is.numeric)),
  perplexity = 15
)
tibble(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y)) +
  geom_point()
```

## Cases vs controls

```{r}
imm_treg_tmem.tsne.in |>
  bind_cols(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y, color = group)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Site of onset

```{r}
imm_treg_tmem.tsne.in |>
  bind_cols(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, site_of_onset),
    by = "sample_id"
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = site_of_onset)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Slow vs fast progressors

```{r}
imm_treg_tmem.tsne.in |>
  bind_cols(tSNE.x = imm_treg_tmem.tsne$Y[, 1], tSNE.y = imm_treg_tmem.tsne$Y[, 2]) |>
  left_join(
    clinical |> select(sample_id, progression_category),
    by = "sample_id"
  ) |>
  filter(group == "ALS") |>
  drop_na(progression_category) |>
  ggplot(aes(tSNE.x, tSNE.y, color = progression_category)) +
  geom_point() +
  theme(legend.position = "bottom")
```


# CD4 cytokine expression

```{r}
imm_cyto_cd4.tsne.in <- imm_cytokines_l.imputed |>
  filter(celltype == "CD4", timepoint == "T0") |>
  select(-celltype, -timepoint)
imm_cyto_cd4.tsne <- Rtsne(imm_cyto_cd4.tsne.in |> select(-sample_id, -group), perplexity = 15)
tibble(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y)) +
  geom_point()
```

## Cases vs controls

```{r}
imm_cyto_cd4.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y, color = group)) +
  geom_point() +
  theme(legend.position = "bottom")
```


## Site of onset

```{r}
imm_cyto_cd4.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, site_of_onset),
    by = "sample_id"
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = site_of_onset)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Slow vs fast progressors

```{r}
imm_cyto_cd4.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd4.tsne$Y[, 1], tSNE.y = imm_cyto_cd4.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, progression_category),
    by = "sample_id"
  ) |>
  drop_na(progression_category) |>
  ggplot(aes(tSNE.x, tSNE.y, color = progression_category)) +
  geom_point() +
  theme(legend.position = "bottom")
```

# CD8 cytokine expression

```{r}
imm_cyto_cd8.tsne.in <- imm_cytokines_l.imputed |>
  filter(celltype == "CD8", timepoint == "T0") |>
  select(-celltype, -timepoint, -ends_with("_tregs"))

imm_cyto_cd8.tsne <- Rtsne(imm_cyto_cd8.tsne.in |> select(-sample_id, -group), perplexity = 15)
tibble(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y)) +
  geom_point()
```

## Cluster analysis

```{r}
imm_cyto_cd8.kmc <- kmeans(
  imm_cyto_cd8.tsne.in |> select(where(is.numeric)),
  centers = 3, nstart = 10
)

imm_cyto_cd8.tsne.in |>
  bind_cols(
    tSNE.x = imm_cyto_cd8.tsne$Y[, 1],
    tSNE.y = imm_cyto_cd8.tsne$Y[, 2],
    KM = factor(imm_cyto_cd8.kmc$cluster)
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = KM)) +
  geom_point() +
  theme(legend.position = "bottom")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(clinical, by = "sample_id") |>
  select(KM, delta_fs, progression_category, site_of_onset) |>
  tbl_summary(by = "KM") |>
  add_p()
```

## Principal component analysis

```{r}
imm_cyto_cd8.pca <- prcomp(
  imm_cyto_cd8.tsne.in |> select(-sample_id, -group),
  center = TRUE, scale. = TRUE)
```

```{r}
tibble(eigenval = imm_cyto_cd8.pca$sdev^2) |>
  mutate(PC = row_number()) |>
  ggplot(aes(PC, eigenval / sum(eigenval))) +
  geom_point() + geom_line() +
  labs(x = "Principal component", y = "Eigenvalue") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal()
```

We can see that the first 5 PCs explain around `r round(sum(imm_cyto_cd8.pca$sdev[1:5]^2) / sum(imm_cyto_cd8.pca$sdev^2) * 100, 1)`% of the variance in the data.

```{r}
imm_cyto_cd8.pca$rotation[, 1:5] |>
  as.data.frame() |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "PC", values_to = "loading") |>
  ggplot(aes(PC, marker, fill = loading)) +
  geom_tile() +
  geom_text(aes(label = round(loading, 2)), size = 3) +
  scale_fill_gradient2(low="blue", mid="white", high="red") +
  labs(x = NULL, y = NULL) +
  theme_minimal()
```

This is how the first 5 PCs are driven by the different cytokine markers:

- **PC1**: driven positively by IL-2, IL-4, GM-CSF and TNF-alpha.
- **PC2**: driven positively by IL-17 and IL-22.
- **PC3**: driven negatively by IFN-gamma, IL-21, CD-161 and TNF-alpha.
- **PC4**: driven positively by CD-161, IL-17 and IL-22, and negatively by IFN-gamma.
- **PC5**: driven negatively by CD-161 and negatively by IL-21 and IL-22.

Let's see how the first 5 PCs are represented among the clusters we identified earlier:

```{r}
imm_cyto_cd8.tsne.in |>
  select(sample_id, group) |>
  bind_cols(KM=imm_cyto_cd8.kmc$cluster) |>
  bind_cols(imm_cyto_cd8.pca$x[,1:5]) |>
  pivot_longer(starts_with("PC"), names_to = "PC", values_to = "value") |>
  ggplot(aes(factor(KM), value, fill=factor(KM))) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("1", "3"), c("1", "2"), c("2", "3")),
    map_signif_level = TRUE,
    textsize = 3,
    size = 0.5,
    y_position = c(8.5, 7, 7)
  ) +
  coord_cartesian(ylim=c(-5, 10)) +
  facet_wrap(~PC) +
  labs(title = "Imputed dataset")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(cd8_group=factor(imm_cyto_cd8.kmc$cluster)) |>
  pivot_longer(
    -c(sample_id, group, cd8_group),
    names_to = "PC", values_to = "value"
  ) |>
  ggplot(aes(cd8_group, value, fill=cd8_group)) +
  geom_boxplot() +
  facet_wrap(~PC, scales = "free") +
  labs(title = "Imputed dataset")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM=imm_cyto_cd8.kmc$cluster) |>
  pivot_longer(-c(sample_id, group, KM), names_to = "marker") |>
  group_by(marker) |>
  reframe(kruskal.test(value ~ KM) |> tidy()) |>
  arrange(p.value)
```


```{r}
imm_cyto_cd8.tsne.in |>
  select(sample_id) |>
  bind_cols(KM=factor(imm_cyto_cd8.kmc$cluster)) |>
  inner_join(imm_cytokines, by = "sample_id") |>
  drop_na(KM, CD8_il21_T0) |>
  ggplot(aes(KM, CD8_il21_T0, fill=KM)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("1", "3"), c("1", "2"), c("2", "3")),
    map_signif_level = TRUE,
    textsize = 3,
    size = 0.5,
    y_position = c(100, 90, 90)
  ) +
  coord_cartesian(ylim=c(0, 110)) +
  labs(title = "Raw dataset")
```

## Cases vs controls

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  ggplot(aes(tSNE.x, tSNE.y, color = group)) +
  geom_point() +
  theme(legend.position = "bottom")
```

## Site of onset

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, site_of_onset),
    by = "sample_id"
  ) |>
  ggplot(aes(tSNE.x, tSNE.y, color = site_of_onset)) +
  geom_point() +
  theme(legend.position = "bottom")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(
    clinical |> select(sample_id, site_of_onset),
    by = "sample_id"
  ) |>
  drop_na(site_of_onset) |>
  mutate(site_of_onset = site_of_onset) |>
  ggplot(aes(KM, fill=site_of_onset)) +
  geom_bar(position="fill")
```

## Slow vs fast progressors

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(tSNE.x = imm_cyto_cd8.tsne$Y[, 1], tSNE.y = imm_cyto_cd8.tsne$Y[, 2]) |>
  filter(group == "ALS") |>
  left_join(
    clinical |> select(sample_id, progression_category),
    by = "sample_id"
  ) |>
  drop_na(progression_category) |>
  ggplot(aes(tSNE.x, tSNE.y, color = progression_category)) +
  geom_point() +
  theme(legend.position = "bottom")
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(
    clinical |> select(sample_id, delta_fs),
    by = "sample_id"
  ) |>
  drop_na(delta_fs) |>
  ggplot(aes(KM, delta_fs, fill=KM)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("1", "3"), c("1", "2"), c("2", "3")),
    map_signif_level = TRUE,
    y_position = c(4, 3.5, 3.5)
  ) +
  coord_cartesian(ylim=c(0, 4.5))
```

```{r}
imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(clinical, by = "sample_id") |>
  ggplot(aes(il21, delta_fs, color=KM)) +
  geom_point()
```

## Patient survival

```{r}
imm_cyto_cd8.surv <- imm_cyto_cd8.tsne.in |>
  bind_cols(KM = factor(imm_cyto_cd8.kmc$cluster)) |>
  left_join(clinical, by = "sample_id")

imm_cyto_cd8.km <- survfit2(
  Surv(disease_duration, vital_status) ~ KM,
  data = imm_cyto_cd8.surv
)

ggsurvfit(imm_cyto_cd8.km) + add_pvalue() + labs(title="Imputed dataset")
```

There seems to be a worse survival associated with CD8-related cluster 2.

```{r}
imm_cyto_cd8_sex.cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM + sex,
  data = imm_cyto_cd8.surv
)

summary(imm_cyto_cd8_sex.cph)
```

```{r}
imm_cyto_cd8_sex_i.cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM * sex,
  data = imm_cyto_cd8.surv
)

summary(imm_cyto_cd8_sex_i.cph)
```

This effect is independent of sex, and there's not an associated interaction.

```{r}
imm_cyto_cd8_onset.cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM + site_of_onset,
  data = imm_cyto_cd8.surv
)

summary(imm_cyto_cd8_onset.cph)
```

```{r}
imm_cyto_cd8_onset_i.cph <- coxph(
  Surv(disease_duration, vital_status) ~ KM * site_of_onset,
  data = imm_cyto_cd8.surv
)

summary(imm_cyto_cd8_onset_i.cph)
```

This effect is also independent of site of onset, and there's not a clear interaction either.

To establish which cytokines among the ones differentially expressed in cluster 2 drive the survival effects, let's analyze each one of them separately:

```{r}
coxph(Surv(disease_duration, vital_status) ~ il21, imm_cyto_cd8.surv)
```

```{r}
coxph(Surv(disease_duration, vital_status) ~ ifn_gamma, imm_cyto_cd8.surv)
```

```{r}
coxph(Surv(disease_duration, vital_status) ~ tnf_alpha, imm_cyto_cd8.surv)
```

```{r}
coxph(Surv(disease_duration, vital_status) ~ gm_csf, imm_cyto_cd8.surv)
```

Among IL-21, IFN-gamma, TNF-alpha and GM-CSF, IL-21 is the only one that seems to have a significant effect on survival.

```{r}
coxph(Surv(disease_duration, vital_status) ~ KM + il21, imm_cyto_cd8.surv)
```

Adjusting for IL-21 makes the effect of CD8-related cluster 2 non-significant, which suggests that IL-21 is the main driver of the survival effect we see in this cluster.

> **Hypothesis**: IL-21 expression in CD8 is the main driver of the survival differences we see in this dataset. Alternatively, it could be inversely-associated with survival if patients with more rapidly progressive ALS or at more advanced stages of disease have higher IL-21 expression by CD8.

Let's test that hypothesis graphically by plotting survival curves of patients above and below the median IL-21:

```{r}
imm_cyto_cd8.surv |>
  mutate(il21_group = if_else(il21 <= median(il21), "IL-21/Low", "IL-21/High")) |>
  with(survfit2(Surv(disease_duration, vital_status) ~ il21_group)) |>
  ggsurvfit() + add_pvalue() + labs(title = "Imputed dataset")
```

We see that patients in the IL-21/High group have a significantly shorter survival time compared to those in the IL-21/Low group (p = 0.03). This suggests that IL-21 is a key factor in determining survival differences in this dataset.

```{r}
imm_cytokines_l |>
  filter(group == "ALS", celltype=="CD8") |>
  drop_na(sample_id, timepoint, il21) |>
  ggplot(aes(timepoint, il21, group=sample_id)) +
  geom_line() +
  coord_cartesian(ylim=c(0, 17)) +
  labs(title = "Raw dataset", x = NULL, y = "CD8 IL-21")
```

```{r}
imm_cytokines.imputed |>
  filter(group == "ALS") |>
  mutate(il21_group = factor(case_when(
    CD8_il21_T0 > median(CD8_il21_T0, na.rm = TRUE) ~ "IL-21/EarlyHigh",
    CD8_il21_T1 > median(CD8_il21_T1, na.rm = TRUE) ~ "IL-21/LateHigh",
    TRUE ~ "IL-21/Low"
  ), levels = c("IL-21/Low", "IL-21/EarlyHigh", "IL-21/LateHigh"))) |>
  left_join(clinical, by = "sample_id") |>
  with(survfit2(Surv(disease_duration, vital_status) ~ il21_group)) |>
  ggsurvfit() + add_pvalue() + labs(title="Imputed dataset")
```

Patients with increased CD8 IL-21 expression at both T0 and T1 show a significantly worse survival compared to those with low IL-21 expression at both timepoints on the imputed dataset (p = 0.039).

```{r}
imm_cytokines_t0 |>
  filter(group == "ALS", timepoint == "T0") |>
  mutate(il21_group = ifelse(
    CD8_il21 <= median(CD8_il21, na.rm = TRUE), "IL-21/Low", "IL-21/High"
  )) |>
  left_join(clinical, by = "sample_id") |>
  with(survfit2(Surv(disease_duration, vital_status) ~ il21_group)) |>
  ggsurvfit() + add_pvalue() + labs(title = "Raw dataset")
```

```{r}
imm_cytokines |>
  filter(group == "ALS") |>
  mutate(il21_group = factor(case_when(
    CD8_il21_T0 > median(CD8_il21_T0, na.rm = TRUE) ~ "IL-21/EarlyHigh",
    CD8_il21_T1 > median(CD8_il21_T1, na.rm = TRUE) ~ "IL-21/LateHigh",
    TRUE ~ "IL-21/Low"
  ), levels = c("IL-21/Low", "IL-21/EarlyHigh", "IL-21/LateHigh"))) |>
  left_join(clinical, by = "sample_id") |>
  with(survfit2(Surv(disease_duration, vital_status) ~ il21_group)) |>
  ggsurvfit() + add_pvalue() + labs(title = "Raw dataset")
```

The same analysis shows a similar trend in the unimputed dataset, but results do not reach statistical significance. This might be due to decreased power from the smaller sample size due to listwise deletion.

```{r}
imm_cd8_il21_t0.surv <- imm_cytokines_t0 |>
  filter(group == "ALS", timepoint == "T0") |>
  mutate(il21_group = ifelse(
    CD8_il21 <= median(CD8_il21, na.rm = TRUE), "C", "E"
  )) |>
  left_join(
    clinical |> select(sample_id, disease_duration, vital_status),
    by = "sample_id"
  )

imm_cd8_il21_t0.pwr <- powerCT(
  formula=Surv(disease_duration, vital_status) ~ il21_group,
  dat=imm_cd8_il21_t0.surv,
  nC=sum(imm_cd8_il21_t0.surv$il21_group == "C", na.rm = TRUE),
  nE=sum(imm_cd8_il21_t0.surv$il21_group == "E", na.rm = TRUE),
  alpha = 0.05,
  RR=(
    imm_cyto_cd8.surv |>
      mutate(
        il21_group = factor(
          if_else(il21 <= median(il21), "IL-21/Low", "IL-21/High"),
          levels = c("IL-21/Low", "IL-21/High")
      )) |>
      with(coxph(Surv(disease_duration, vital_status) ~ il21_group)) |>
      coef() |> exp() |> as.numeric()
  )
)
```

From the code above, we have an estimated power of only `r round(imm_cd8_il21_t0.pwr$power * 100, 1)`% to detect a difference as significant as that seen in the imputed dataset with alpha=0.05.